A Handbook for Tenured Track Faculty in Computer Science in the US

### 2024-03-22 faculty.tex
updates

\documentclass[oneside,11pt]{memoir}
\chapterstyle{pedersen}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[margin=1.5in]{geometry}
\usepackage{soul}
% \usepackage[small,compact]{titlesec} %very powerful
\usepackage[most]{tcolorbox}
\setsecnumdepth{subsection}
\setcounter{tocdepth}{3}
\usepackage{enumitem}
\usepackage{epigraph}
\usepackage{cite}
\usepackage{caption}
\captionsetup{font=small}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{wrapfig}
\setlength\intextsep{0pt} % remove extra space above and below in-line float
\usepackage{hyperref}
\hypersetup{
  colorlinks,
  citecolor=black,
  filecolor=black,
  linkcolor=blue,
  urlcolor=blue,
}
\usepackage{booktabs}
\usepackage{xcolor}
\usepackage[makeroom]{cancel}

@@

\title{A Handbook for Tenured Track Faculty in Computer Science in the US}

\author{\href{https://nguyenthanhvuh.github.io}{ThanhVu (Vu) Nguyen}\\{\small George Mason University, Dept. of Computer Science}}

\makeatletter
\def\maketitle{%
  \null
  \thispagestyle{empty}%
  \vfill
  \begin{center}\leavevmode
    \normalfont
    {\LARGE\raggedright \textbf{\@title}\par}%
    \vfill%
    % \hrulefill\par
    {\Large \@author\par}%
    \vfill%
    {\large\raggedleft \@date\par}%
  \end{center}%
  \vfill
  \null
  \cleardoublepage
}
\makeatother

\begin{document}
\maketitle
\frontmatter

@@

Chapter 3: Funding
Let’s talk about what, for many professors, is the least favorite part of the job: making sure you
have enough money to keep the lights on in the lab. Essentially, a professor applies for grants
(money from the government, industry, or other charitable foundations) by writing a proposal as
the Principal Investigator (PI). The funding agency reviews the proposal. If they decide to fund it,
they send the money to your university. The university then takes its cut of the funding
(essentially to pay for common services like buildings, admins, etc) and gives you the rest. You
then spend it on your summer salary, the salaries of your students, equipment, travel,
registration costs and other expenses associated with your research.
There are some nuances about the different types of funding, and how you can spend different
kinds of funding. So let us get into the details. Unfortunately, these details tend to vary a lot by
country, so everything I’m talking about in this chapter applies only to the US.
3.1 An example budget
Let me walk you through an example budget. Let us say you are writing a proposal for \$100,000
USD (100K USD). The budget will talk about how you plan to spend this money. The following
table (Table 1) shows the example budget.
Budget
Graduate Research Assistant - 1
Total Salary 40,000
Fringe Benefits 12,000
Total Salary and Fringe Benefits 52,000
Total Materials \& Supplies 0
International travel to a conference 2500
Domestic travel to a conference 1481
Total Travel Costs 3,981
Tuition for Fall or Spring 4,746
Tuition for Summer 1,778
GRA Tuition for 1 year 11,270
Total Direct Costs 67,251
Modified Total Direct Cost (MTDC) 55,981
Indirect Costs 58.5% of MTDC 32,749
Total Cost (Direct + indirect costs) 100,000
37
I will explain the various parts of this example budget. I will be talking about terms such as
direct costs and indirect costs as defined by the University of Texas – I believe it is similar at
most other US universities, but there might be differences. Please check with your university to
make sure.
Fringe. Let us say you want to support a student on this grant. In addition to paying for the
student’s salary, you also have to pay for non-wage expenses: things like health insurance, social
security, and medicare. This is usually expressed as a percentage of the student’s salary. At the
University of Texas at Austin, this was 30% in 2022 (For a comparison point, it was 32% at
Princeton). The stipend of a graduate research assistant (GRA) at UT is about \$30K USD for the
fall and spring semesters ($40K USD including summer). So salary and fringe for a GRA for a
year is \$52K.
You also have to pay the tuition of the GRA, and at UT, this is about \$11.2K USD for a year. To
keep things simple, I am using a single number for the tuition, but this depends on factors such
as whether you are in-state or out-of-state, your progress in the PhD program (if you are
all-but-dissertation, your tuition is lower), and whether you are a US citizen or not. It also
depends on the fraction of time spent as GRA/TA: if this drops below 20 hours a week
(considered full-time for students), tuition might have to be charged at the out-of-state rate. If
the student is only doing 5 hours a week as GRA, it might also impact how much of their tuition
can be covered by the grant; at 5 hours a week, the student might not be eligible for insurance. I
highly recommend talking to your admin if you are accounting for less than 20 hours a week for
your student on the project. Some schools also calculate tuition as a % of the stipend.
So the total cost for supporting a GRA for one year is \$63.2K USD. While you are paying \$63.2K
for each student for a year, the student's take-home pay is only the stipend part (minus taxes
that are withheld): \$35.2K assuming a 12% tax rate.
Note that fringe also applies to you (the PI) when you pay yourself out of your grants for your
summer months.
Direct costs. The direct cost part of your budget includes salaries, fringe benefits, travel, and
equipment and supplies that contribute directly towards performing the proposed research.
Direct costs are exclusive to each proposal: a piece of equipment can only be a direct cost on
one proposal, and a GRA can be assigned only to one proposal at a given time. In our example,
let us assume the only direct costs are supporting the student and paying for the student’s
travel (including registration, transportation, lodging, and food) for one international and one
domestic conference.
Modified Total Direct Costs (MTDC). MTDC is a portion of the total direct costs used for
calculating the indirect costs. It is the direct costs minus tuition, fellowships, equipment, and
capital expenditures. Travel also counts as MTDC. In this budget, we are not spending anything
38
on materials and supplies in this budget, but if we did, it would be part of MTDC. Check with
your university to determine exactly what is counted as MTDC – this can be surprisingly tricky.
For example, you might expect computers to not count towards MTDC as it is “equipment”, but it
may be counted as a “supply” if the university is unable to amortize its cost. In the example
above, total direct costs is \$67.2K USD (63.2K for one GRA plus 4K for travel), but MTDC is only
$56K (leaving out the \$11.2K for tuition).
Indirect costs. The university takes a portion of all the money you receive, to pay for centralized
functions such as building costs, utilities, admin, and libraries. This is termed as the “indirect”
cost, usually expressed as a percentage of the MTDC. At most universities in the US, indirect
costs are more than 50% of the MTDC. At UT, it is 58.5% (Mike Freedman reports it is 63% at
Princeton). Thus, in our budget, we would have .585 * 56K = \$32.7K added as indirect costs. As
a result, it would cost 67.2 + 32.7 = \$100K USD for supporting a single GRA and some travel on
this grant. Note that even though the 58.5% number is scary, you are only paying 33% of your
budget for indirect costs; in other words, you don’t lose 58.5% of the total budget to indirect
costs.
As you can see from the preceding table, the \$100K USD budget only covers one student for one
year. At some universities, a GRA costs more, requiring more funding.
This should give you a picture of how expensive it is to maintain large groups! If you had ten
students, you would need to pull in a million dollars each year to fund your students.
It is useful to think of your budget in terms of student-years. \$400K is simply two students for
two years. A dissertation, roughly five student years, is \$500K. If you get the NSF CAREER grant,
you would typically be able to support one dissertation. This helps determine the scope of your
proposal: if you need five students to work on it for two years, a grant that provides \$250K will
not be a good match.
Summer months in your budget. PIs at most R1 universities in the US are expected to fund
themselves via grants in the summer. When you are drawing up your budget, you are expected
to estimate how much of your summer months you will dedicate to this project. This is taken as
an indication of PI commitment at the funding agency – you don’t want to write a budget where
your estimated time involvement is zero. So if you have one NSF grant where you estimate your
time commitment to be one month, you can only put in one more month for another NSF grant
proposal (since NSF limits total PI compensation to be two months per year across all NSF
grants).
3.2 Types of grants
Broadly speaking, there are four different kinds of grants or funding, each with different
restrictions on how you can use the money. The differences come down to what you can spend
the money on, how long you can keep the money, and what the overhead is.
39
We can term the four kinds of funding as grants, contracts, university funding, and gift funding.
Let me go over each type of funding.
Grants. The National Science Foundation (NSF), the Defense Advanced Research Projects
Agency (DARPA), the Army, the Navy, and the Air Force all fund research related to computer
science. These grants are competitive – NSF has a 25% acceptance rate (though this varies by
the directorate), and I imagine the other programs are similar.
This money can typically only be used for funding students, yourself, travel, and equipment
purchases that you had already put in the budget. For example, if you suddenly want to
purchase equipment not in the original budget, it needs to be approved by the Office of
Sponsored Research at your university (which makes sure you are spending the money in the
right way). You cannot use government funding to pay for things like a meal with your students
during the semester (meals during travel is allowed).
Government funding has strict timelines by which the money has to be spent. Typically, the
money is released on a yearly basis. You have to show that you have used the previous
installment to get the next round of funding. This sometimes results in professors making last
minute purchases to exhaust their grant!
In some cases, you can ask your funding program manager for a no-cost extension to your
grant. This typically means that you don’t need extra money, but you want extra time to use the
money. This is usually considered on a case-by-case basis, and I would not recommend
counting on getting a no-cost extension.
Indirect costs are levied on government funding (more generally, on all funding that goes
through the Office of Sponsored Research). The example budget I showed in the previous
section assumes funding is via sponsored research.
Grant proposals typically have reporting requirements (for example, you have to submit a report
once a year for an NSF grant). However, grants typically do not expect any concrete deliverables
to be provided at the end of the funding period. Essentially, the grant allows you to carry out the
proposed research: the act of publishing itself is the desired outcome. As a result, grants have a
strong alignment between the PIs and the funding agency: they both just want to get research
done.
Contracts. The second kind of funding is via contracts. Contract funding comes with an explicit
contract you must fulfill to keep and obtain more funding. For example, there may be quarterly
or monthly reports, or on-site visits by the sponsor. Contract funding typically also has
deliverables that are expected at regular intervals during the funding period. One example of
contract funding is DARPA funding, which comes with a number of deliverables and reporting
requirements. Contract funding is less aligned with the PIs than grant funding: apart from doing
the core research, the PI’s students will have to spend time on creating the deliverables, and the
PI herself has to spend a significant amount of time on reports and meetings.
40
NSF vs DARPA. NSF is a good example of grant funding, while DARPA is a good example of
contract funding. Both these agencies fund a lot of computer science research. DARPA requires
deliverables and reports to be produced regularly during the grant duration. NSF only requires a
report to be submitted for each year of the grant duration. You can apply for NSF funds as
someone on a visa, while DARPA and other similar government funding programs might require
you to be a US citizen or a permanent resident.
University funding. Your university, or your college, or your department may choose to give you
funding. Typically when you start, you get a “startup” package that allows you to kickstart your
research group. University funding is “internal” funding in a way, and as such has fewer
restrictions than government funding. But this really depends on the specific university. I know
universities where startup funding cannot be used for covering summer salary of PIs. Apart
from startup funding, your university might have internal competitions to award “seed” funding
(typically under 50K) that you can then use to get preliminary results and later apply for
traditional grants.
University funding is also associated with timelines. For example, startup funding usually has to
be spent before you get tenure. However, university funding deadlines might be a little more
flexible than grant or contract funding deadlines.
University funding typically does not have associated indirect costs (as it is already university
money). You still have to pay fringe support for your students and your own summer salary.
An interesting sub-category of university funding are Research Centers. A research center is
funded by sponsors to carry out a certain mission over a span of around 5 years. The research
center has a program director who is responsible for allocating the funding. PIs at the university
apply for funding from the research center (a sort of internal grant), and are allocated funding.
Obtaining funding in this manner is easier than competing for an external grant. Particularly for
assistant professors who are just starting, research centers can provide seed funding before the
first grants are obtained. Check out the ADA center at Michigan for a good example.
Gift funding or unrestricted research funding. Companies or charitable foundations may
choose to fund your research in the form of “gifts” or unrestricted research grants. This is the
best kind of funding, for three reasons: indirect costs are typically not levied on these gifts, there
is no associated timeline by when you should spend them, and you can use them for a broader
range of things than university and grant/contract funding.
Indirect costs are not levied because the gift-giver can specify that it cannot be used for indirect
costs (the university has to honor this). Since it is an unrestricted gift, there are no limits on
what you can spend them on, or by when you should spend them. This doesn’t mean you can
spend the money on personal expenses. All expenses still have to be approved by the university,
and should be used for research-related expenses. But things like taking your research group
out for a meal can be expensed during gift funding (but not government funding).
41
Indirect costs not being levied can make a huge difference. For example, in our budget from
earlier, if you were using gift funding to fund the student, you don’t have to pay indirect costs,
and thus get over \$30K back. Considering a student costs \$63K, you get to sponsor the student
for 1.5 years rather than 1 year!
How should you use your funding? Let us say that you are lucky enough to have all four kinds of
funding. And you have an expense that could be assigned to any of these funding (note that this
is often not the case: you can’t use a grant on topic X to support a student working on Y if X and
Y are not closely related). What should you use?
The typical rule is “Use the most restricted funding that will expire the earliest”. Thus, you should
always use the appropriate grant/contract funding first, then your university funding, and finally
gift funding. Gift funding is the most flexible, so use it only when you have no other option.
Always spend your gift funding on expenses that will have overhead (if funded from grants),
such as students. Do not spend it on things like equipment (be careful to ask what counts as
equipment at your university) which do not have overhead even when expensed to grants.
Writing a grant with other PIs. While you could write smaller grants as the sole PI, typically for
larger grants, you need to collaborate with other PIs and submit as a team. For example, you
cannot submit NSF Large grants as a solo PI. One thing to remember though: all the funding is
shared between the PIs. For example, imagine a grant for one million USD. It sounds amazing!
However, if the grant was awarded to a team of two PIs, and the grant duration is five years,
each PI can only support one student on it for the grant duration (100K * 2 * 5). Similarly, if a
grant of 10 million USD was won by a team of five PIs, each PI can only support one student for
four years (100K * 5 * 4). So in the end, the share of each PI is similar or lower than that of an
NSF CAREER.
This is not to say you should not collaborate with other PIs to write grants. Collaboration often
brings new ideas and research directions that are unlikely or impossible when working alone.
And bigger grants often provide prestige: for example, landing an NSF Large grant is significant
and puts you on the map. But it is useful to calculate how much you expect to receive if a grant
proposal is funded.
3.3 The grant application process
So how does one go about submitting a grant? There are several steps in the process:
1. The funding agency first puts out a call for proposals. For example, check out the NSF
call for CRII grants, a startup grant meant for new PIs who have not yet received a grant.
Note that PIs at many R1 universities are not eligible for CRII. This was posted on May 18,
2022. Each call has a solicitation page with details about the program and how the
proposal should be structured – you should read this carefully.
42
2. In some cases, there are webinars where the program managers explain the grant
opportunity. CRII had such a webinar on Jul 13 2022. I highly recommend attending
these webinars if it is your first time submitting for a particular grant opportunity.
3. In some cases, the funding agency might want you to submit a preliminary proposal.
They will review the preliminary proposals and inform you if they would like you to go
ahead with the full proposal. This is not the case for CRII.
4. There will be a due date by which you need to submit the proposal. For the CRII, this is
Sep 19 2022. The proposal will have a number of parts, and the solicitation will clearly lay
out what the proposal should look like.
5. You write the proposal, and sometime before you submit, you send it to the Office of
Sponsored Research (OSP or OSR) at your university. They have to approve the proposal
before you can submit, and they may request to see your proposal a week or 10 days
before you submit. Please make sure you get a draft to the OSP on time!
6. Finally, you submit the proposal, and you wait. To be exact, your admin/OSP submits for
you, so you want to let them know when you are done. You and your admin/OSP will be
submitting a lot of proposals together; make sure you get to know the folks doing this. I
only have experience with NSF proposals, but these usually take 6-8 months to be
reviewed. In some cases, it might take even longer.
7. If you are lucky enough to get your proposal funded, you might get an email or a call from
the NSF program manager. They might ask you to slightly modify your budget. For
example, I was asked to reduce my NSF CAREER budget (and appropriately reduce the
scope of my tasks as well). You will be asked to provide an abstract that will be displayed
on the NSF website. For example, check out the abstract for my CAREER award.
8. Finally, you will get an official email that your grant has been awarded, it will become
public on the NSF website, and you will get the first installment delivered to your
university. You can then begin spending it in accordance with what you proposed.
9. In the unlucky case that your proposal is not funded, you will get back three or more

>>>>

\documentclass[oneside,11pt,dvipsnames]{book}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[margin=1.5in]{geometry}
\usepackage{soul}
% \usepackage[small,compact]{titlesec} %very powerful
\usepackage[most]{tcolorbox}
% \setsecnumdepth{subsection}
% \setcounter{tocdepth}{3}
\usepackage{enumitem}
\usepackage{epigraph}
\usepackage{cite}
\usepackage{caption}
\captionsetup{font=small}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{wrapfig}
\setlength\intextsep{0pt} % remove extra space above and below in-line float
\usepackage{hyperref}
\hypersetup{
  colorlinks,
  citecolor=black,
  filecolor=black,
  linkcolor=blue,
  urlcolor=blue,
}
\usepackage{booktabs}

\usepackage{tikz}
\usetikzlibrary{calc}
\usepackage{xcolor}

\usepackage{anyfontsize}
\usepackage{sectsty}

\usepackage[makeroom]{cancel}

@@

\begin{document}

\pagestyle{empty}
\begin{tikzpicture}[overlay,remember picture]

    % Background color
    \fill[
    black!2]
    (current page.south west) rectangle (current page.north east);

    % Rectangles
    \shade[
    left color=Dandelion, 
    right color=Dandelion!40,
    transform canvas ={rotate around ={45:($(current page.north west)+(0,-6)$)}}] 
    ($(current page.north west)+(0,-6)$) rectangle ++(9,1.5);

    \shade[
    left color=lightgray,
    right color=lightgray!50,
    rounded corners=0.75cm,
    transform canvas ={rotate around ={45:($(current page.north west)+(.5,-10)$)}}]
    ($(current page.north west)+(0.5,-10)$) rectangle ++(15,1.5);

    \shade[
    left color=lightgray,
    rounded corners=0.3cm,
    transform canvas ={rotate around ={45:($(current page.north west)+(.5,-10)$)}}] ($(current page.north west)+(1.5,-9.55)$) rectangle ++(7,.6);

    \shade[
    left color=orange!80,
    right color=orange!60,
    rounded corners=0.4cm,
    transform canvas ={rotate around ={45:($(current page.north)+(-1.5,-3)$)}}]
    ($(current page.north)+(-1.5,-3)$) rectangle ++(9,0.8);

    \shade[
    left color=red!80,
    right color=red!80,
    rounded corners=0.9cm,
    transform canvas ={rotate around ={45:($(current page.north)+(-3,-8)$)}}] ($(current page.north)+(-3,-8)$) rectangle ++(15,1.8);

    \shade[
    left color=orange,
    right color=Dandelion,
    rounded corners=0.9cm,
    transform canvas ={rotate around ={45:($(current page.north west)+(4,-15.5)$)}}]
    ($(current page.north west)+(4,-15.5)$) rectangle ++(30,1.8);

    \shade[
    left color=RoyalBlue,
    right color=Emerald,
    rounded corners=0.75cm,
    transform canvas ={rotate around ={45:($(current page.north west)+(13,-10)$)}}]
    ($(current page.north west)+(13,-10)$) rectangle ++(15,1.5);

    \shade[
    left color=ForestGreen,
    rounded corners=0.3cm,
    transform canvas ={rotate around ={45:($(current page.north west)+(18,-8)$)}}]
    ($(current page.north west)+(18,-8)$) rectangle ++(15,0.6);

    \shade[
    left color=ForestGreen,
    rounded corners=0.4cm,
    transform canvas ={rotate around ={45:($(current page.north west)+(19,-5.65)$)}}]
    ($(current page.north west)+(19,-5.65)$) rectangle ++(15,0.8);

    \shade[
    left color=OrangeRed,
    right color=red!80,
    rounded corners=0.6cm,
    transform canvas ={rotate around ={45:($(current page.north west)+(20,-9)$)}}] 
    ($(current page.north west)+(20,-9)$) rectangle ++(14,1.2);

    % Year
    \draw[ultra thick,gray]
    ($(current page.center)+(5,2)$) -- ++(0,-3cm) 
    node[
    midway,
    left=0.25cm,
    text width=5cm,
    align=right,
    black!75
    ]
    {
    {\fontsize{25}{30} \selectfont \bf Computer\\[8pt] Science}
    } 
    node[
    midway,
    right=0.25cm,
    text width=6cm,
    align=left,
    ForestGreen]
    {
    {\fontsize{48}{50} \selectfont PhD}
    };

    % Title
    \node[align=center] at ($(current page.center)+(0,-5)$) 
    {
    {\fontsize{24}{24} \selectfont {{Demystifying the Computer Science}}} \\[0.15in]
    {\fontsize{24}{24} \selectfont {{PhD Admission in the US}}} \\[0.2in]    
    {\fontsize{18}{18} \selectfont {{A Handbook for International and Domestic Students}}} \\[0.5in]

    {\fontsize{16}{19.2} \selectfont \textcolor{ForestGreen}{ \bf ThanhVu (Vu) Nguyen}}\\[0.1in]
    George Mason University, Dept. of Computer Science\\[0.1in]
    \today{} (latest version available on  \href{https://github.com/nguyenthanhvuh/phd-cs-us}{Github})
    };
    \end{tikzpicture}

@@

% Chapter 3: Funding
% Let’s talk about what, for many professors, is the least favorite part of the job: making sure you
% have enough money to keep the lights on in the lab. Essentially, a professor applies for grants
% (money from the government, industry, or other charitable foundations) by writing a proposal as
% the Principal Investigator (PI). The funding agency reviews the proposal. If they decide to fund it,
% they send the money to your university. The university then takes its cut of the funding
% (essentially to pay for common services like buildings, admins, etc) and gives you the rest. You
% then spend it on your summer salary, the salaries of your students, equipment, travel,
% registration costs and other expenses associated with your research.
% There are some nuances about the different types of funding, and how you can spend different
% kinds of funding. So let us get into the details. Unfortunately, these details tend to vary a lot by
% country, so everything I’m talking about in this chapter applies only to the US.
% 3.1 An example budget
% Let me walk you through an example budget. Let us say you are writing a proposal for \$100,000
% USD (100K USD). The budget will talk about how you plan to spend this money. The following
% table (Table 1) shows the example budget.
% Budget
% Graduate Research Assistant - 1
% Total Salary 40,000
% Fringe Benefits 12,000
% Total Salary and Fringe Benefits 52,000
% Total Materials \& Supplies 0
% International travel to a conference 2500
% Domestic travel to a conference 1481
% Total Travel Costs 3,981
% Tuition for Fall or Spring 4,746
% Tuition for Summer 1,778
% GRA Tuition for 1 year 11,270
% Total Direct Costs 67,251
% Modified Total Direct Cost (MTDC) 55,981
% Indirect Costs 58.5% of MTDC 32,749
% Total Cost (Direct + indirect costs) 100,000
% 37
% I will explain the various parts of this example budget. I will be talking about terms such as
% direct costs and indirect costs as defined by the University of Texas – I believe it is similar at
% most other US universities, but there might be differences. Please check with your university to
% make sure.
% Fringe. Let us say you want to support a student on this grant. In addition to paying for the
% student’s salary, you also have to pay for non-wage expenses: things like health insurance, social
% security, and medicare. This is usually expressed as a percentage of the student’s salary. At the
% University of Texas at Austin, this was 30% in 2022 (For a comparison point, it was 32% at
% Princeton). The stipend of a graduate research assistant (GRA) at UT is about \$30K USD for the
% fall and spring semesters ($40K USD including summer). So salary and fringe for a GRA for a
% year is \$52K.
% You also have to pay the tuition of the GRA, and at UT, this is about \$11.2K USD for a year. To
% keep things simple, I am using a single number for the tuition, but this depends on factors such
% as whether you are in-state or out-of-state, your progress in the PhD program (if you are
% all-but-dissertation, your tuition is lower), and whether you are a US citizen or not. It also
% depends on the fraction of time spent as GRA/TA: if this drops below 20 hours a week
% (considered full-time for students), tuition might have to be charged at the out-of-state rate. If
% the student is only doing 5 hours a week as GRA, it might also impact how much of their tuition
% can be covered by the grant; at 5 hours a week, the student might not be eligible for insurance. I
% highly recommend talking to your admin if you are accounting for less than 20 hours a week for
% your student on the project. Some schools also calculate tuition as a % of the stipend.
% So the total cost for supporting a GRA for one year is \$63.2K USD. While you are paying \$63.2K
% for each student for a year, the student's take-home pay is only the stipend part (minus taxes
% that are withheld): \$35.2K assuming a 12% tax rate.
% Note that fringe also applies to you (the PI) when you pay yourself out of your grants for your
% summer months.
% Direct costs. The direct cost part of your budget includes salaries, fringe benefits, travel, and
% equipment and supplies that contribute directly towards performing the proposed research.
% Direct costs are exclusive to each proposal: a piece of equipment can only be a direct cost on
% one proposal, and a GRA can be assigned only to one proposal at a given time. In our example,
% let us assume the only direct costs are supporting the student and paying for the student’s
% travel (including registration, transportation, lodging, and food) for one international and one
% domestic conference.
% Modified Total Direct Costs (MTDC). MTDC is a portion of the total direct costs used for
% calculating the indirect costs. It is the direct costs minus tuition, fellowships, equipment, and
% capital expenditures. Travel also counts as MTDC. In this budget, we are not spending anything
% 38
% on materials and supplies in this budget, but if we did, it would be part of MTDC. Check with
% your university to determine exactly what is counted as MTDC – this can be surprisingly tricky.
% For example, you might expect computers to not count towards MTDC as it is “equipment”, but it
% may be counted as a “supply” if the university is unable to amortize its cost. In the example
% above, total direct costs is \$67.2K USD (63.2K for one GRA plus 4K for travel), but MTDC is only
% $56K (leaving out the \$11.2K for tuition).
% Indirect costs. The university takes a portion of all the money you receive, to pay for centralized
% functions such as building costs, utilities, admin, and libraries. This is termed as the “indirect”
% cost, usually expressed as a percentage of the MTDC. At most universities in the US, indirect
% costs are more than 50% of the MTDC. At UT, it is 58.5% (Mike Freedman reports it is 63% at
% Princeton). Thus, in our budget, we would have .585 * 56K = \$32.7K added as indirect costs. As
% a result, it would cost 67.2 + 32.7 = \$100K USD for supporting a single GRA and some travel on
% this grant. Note that even though the 58.5% number is scary, you are only paying 33% of your
% budget for indirect costs; in other words, you don’t lose 58.5% of the total budget to indirect
% costs.
% As you can see from the preceding table, the \$100K USD budget only covers one student for one
% year. At some universities, a GRA costs more, requiring more funding.
% This should give you a picture of how expensive it is to maintain large groups! If you had ten
% students, you would need to pull in a million dollars each year to fund your students.
% It is useful to think of your budget in terms of student-years. \$400K is simply two students for
% two years. A dissertation, roughly five student years, is \$500K. If you get the NSF CAREER grant,
% you would typically be able to support one dissertation. This helps determine the scope of your
% proposal: if you need five students to work on it for two years, a grant that provides \$250K will
% not be a good match.
% Summer months in your budget. PIs at most R1 universities in the US are expected to fund
% themselves via grants in the summer. When you are drawing up your budget, you are expected
% to estimate how much of your summer months you will dedicate to this project. This is taken as
% an indication of PI commitment at the funding agency – you don’t want to write a budget where
% your estimated time involvement is zero. So if you have one NSF grant where you estimate your
% time commitment to be one month, you can only put in one more month for another NSF grant
% proposal (since NSF limits total PI compensation to be two months per year across all NSF
% grants).
% 3.2 Types of grants
% Broadly speaking, there are four different kinds of grants or funding, each with different
% restrictions on how you can use the money. The differences come down to what you can spend
% the money on, how long you can keep the money, and what the overhead is.
% 39
% We can term the four kinds of funding as grants, contracts, university funding, and gift funding.
% Let me go over each type of funding.
% Grants. The National Science Foundation (NSF), the Defense Advanced Research Projects
% Agency (DARPA), the Army, the Navy, and the Air Force all fund research related to computer
% science. These grants are competitive – NSF has a 25% acceptance rate (though this varies by
% the directorate), and I imagine the other programs are similar.
% This money can typically only be used for funding students, yourself, travel, and equipment
% purchases that you had already put in the budget. For example, if you suddenly want to
% purchase equipment not in the original budget, it needs to be approved by the Office of
% Sponsored Research at your university (which makes sure you are spending the money in the
% right way). You cannot use government funding to pay for things like a meal with your students
% during the semester (meals during travel is allowed).
% Government funding has strict timelines by which the money has to be spent. Typically, the
% money is released on a yearly basis. You have to show that you have used the previous
% installment to get the next round of funding. This sometimes results in professors making last
% minute purchases to exhaust their grant!
% In some cases, you can ask your funding program manager for a no-cost extension to your
% grant. This typically means that you don’t need extra money, but you want extra time to use the
% money. This is usually considered on a case-by-case basis, and I would not recommend
% counting on getting a no-cost extension.
% Indirect costs are levied on government funding (more generally, on all funding that goes
% through the Office of Sponsored Research). The example budget I showed in the previous
% section assumes funding is via sponsored research.
% Grant proposals typically have reporting requirements (for example, you have to submit a report
% once a year for an NSF grant). However, grants typically do not expect any concrete deliverables
% to be provided at the end of the funding period. Essentially, the grant allows you to carry out the
% proposed research: the act of publishing itself is the desired outcome. As a result, grants have a
% strong alignment between the PIs and the funding agency: they both just want to get research
% done.
% Contracts. The second kind of funding is via contracts. Contract funding comes with an explicit
% contract you must fulfill to keep and obtain more funding. For example, there may be quarterly
% or monthly reports, or on-site visits by the sponsor. Contract funding typically also has
% deliverables that are expected at regular intervals during the funding period. One example of
% contract funding is DARPA funding, which comes with a number of deliverables and reporting
% requirements. Contract funding is less aligned with the PIs than grant funding: apart from doing
% the core research, the PI’s students will have to spend time on creating the deliverables, and the
% PI herself has to spend a significant amount of time on reports and meetings.
% 40
% NSF vs DARPA. NSF is a good example of grant funding, while DARPA is a good example of
% contract funding. Both these agencies fund a lot of computer science research. DARPA requires
% deliverables and reports to be produced regularly during the grant duration. NSF only requires a
% report to be submitted for each year of the grant duration. You can apply for NSF funds as
% someone on a visa, while DARPA and other similar government funding programs might require
% you to be a US citizen or a permanent resident.
% University funding. Your university, or your college, or your department may choose to give you
% funding. Typically when you start, you get a “startup” package that allows you to kickstart your
% research group. University funding is “internal” funding in a way, and as such has fewer
% restrictions than government funding. But this really depends on the specific university. I know
% universities where startup funding cannot be used for covering summer salary of PIs. Apart
% from startup funding, your university might have internal competitions to award “seed” funding
% (typically under 50K) that you can then use to get preliminary results and later apply for
% traditional grants.
% University funding is also associated with timelines. For example, startup funding usually has to
% be spent before you get tenure. However, university funding deadlines might be a little more
% flexible than grant or contract funding deadlines.
% University funding typically does not have associated indirect costs (as it is already university
% money). You still have to pay fringe support for your students and your own summer salary.
% An interesting sub-category of university funding are Research Centers. A research center is
% funded by sponsors to carry out a certain mission over a span of around 5 years. The research
% center has a program director who is responsible for allocating the funding. PIs at the university
% apply for funding from the research center (a sort of internal grant), and are allocated funding.
% Obtaining funding in this manner is easier than competing for an external grant. Particularly for
% assistant professors who are just starting, research centers can provide seed funding before the
% first grants are obtained. Check out the ADA center at Michigan for a good example.
% Gift funding or unrestricted research funding. Companies or charitable foundations may
% choose to fund your research in the form of “gifts” or unrestricted research grants. This is the
% best kind of funding, for three reasons: indirect costs are typically not levied on these gifts, there
% is no associated timeline by when you should spend them, and you can use them for a broader
% range of things than university and grant/contract funding.
% Indirect costs are not levied because the gift-giver can specify that it cannot be used for indirect
% costs (the university has to honor this). Since it is an unrestricted gift, there are no limits on
% what you can spend them on, or by when you should spend them. This doesn’t mean you can
% spend the money on personal expenses. All expenses still have to be approved by the university,
% and should be used for research-related expenses. But things like taking your research group
% out for a meal can be expensed during gift funding (but not government funding).
% 41
% Indirect costs not being levied can make a huge difference. For example, in our budget from
% earlier, if you were using gift funding to fund the student, you don’t have to pay indirect costs,
% and thus get over \$30K back. Considering a student costs \$63K, you get to sponsor the student
% for 1.5 years rather than 1 year!
% How should you use your funding? Let us say that you are lucky enough to have all four kinds of
% funding. And you have an expense that could be assigned to any of these funding (note that this
% is often not the case: you can’t use a grant on topic X to support a student working on Y if X and
% Y are not closely related). What should you use?
% The typical rule is “Use the most restricted funding that will expire the earliest”. Thus, you should
% always use the appropriate grant/contract funding first, then your university funding, and finally
% gift funding. Gift funding is the most flexible, so use it only when you have no other option.
% Always spend your gift funding on expenses that will have overhead (if funded from grants),
% such as students. Do not spend it on things like equipment (be careful to ask what counts as
% equipment at your university) which do not have overhead even when expensed to grants.
% Writing a grant with other PIs. While you could write smaller grants as the sole PI, typically for
% larger grants, you need to collaborate with other PIs and submit as a team. For example, you
% cannot submit NSF Large grants as a solo PI. One thing to remember though: all the funding is
% shared between the PIs. For example, imagine a grant for one million USD. It sounds amazing!
% However, if the grant was awarded to a team of two PIs, and the grant duration is five years,
% each PI can only support one student on it for the grant duration (100K * 2 * 5). Similarly, if a
% grant of 10 million USD was won by a team of five PIs, each PI can only support one student for
% four years (100K * 5 * 4). So in the end, the share of each PI is similar or lower than that of an
% NSF CAREER.
% This is not to say you should not collaborate with other PIs to write grants. Collaboration often
% brings new ideas and research directions that are unlikely or impossible when working alone.
% And bigger grants often provide prestige: for example, landing an NSF Large grant is significant
% and puts you on the map. But it is useful to calculate how much you expect to receive if a grant
% proposal is funded.
% 3.3 The grant application process
% So how does one go about submitting a grant? There are several steps in the process:
% 1. The funding agency first puts out a call for proposals. For example, check out the NSF
% call for CRII grants, a startup grant meant for new PIs who have not yet received a grant.
% Note that PIs at many R1 universities are not eligible for CRII. This was posted on May 18,
% 2022. Each call has a solicitation page with details about the program and how the
% proposal should be structured – you should read this carefully.
% 42
% 2. In some cases, there are webinars where the program managers explain the grant
% opportunity. CRII had such a webinar on Jul 13 2022. I highly recommend attending
% these webinars if it is your first time submitting for a particular grant opportunity.
% 3. In some cases, the funding agency might want you to submit a preliminary proposal.
% They will review the preliminary proposals and inform you if they would like you to go
% ahead with the full proposal. This is not the case for CRII.
% 4. There will be a due date by which you need to submit the proposal. For the CRII, this is
% Sep 19 2022. The proposal will have a number of parts, and the solicitation will clearly lay
% out what the proposal should look like.
% 5. You write the proposal, and sometime before you submit, you send it to the Office of
% Sponsored Research (OSP or OSR) at your university. They have to approve the proposal
% before you can submit, and they may request to see your proposal a week or 10 days
% before you submit. Please make sure you get a draft to the OSP on time!
% 6. Finally, you submit the proposal, and you wait. To be exact, your admin/OSP submits for
% you, so you want to let them know when you are done. You and your admin/OSP will be
% submitting a lot of proposals together; make sure you get to know the folks doing this. I
% only have experience with NSF proposals, but these usually take 6-8 months to be
% reviewed. In some cases, it might take even longer.
